### 2016-10-31 ms/ms/ms.tex
added figures to ms

\documentclass[12pt]{article} 
\usepackage[pdftex]{graphicx}
\usepackage{natbib} 
\usepackage{color}
\usepackage{amsmath} 
\usepackage{amssymb} 
\usepackage{verbatim}
\usepackage{mathpazo} 
\usepackage{setspace}
\usepackage{multirow}
\usepackage{fullpage}
\usepackage{lscape}
\usepackage{fancyhdr}
\usepackage[normalem]{ulem} 
\usepackage{hyperref}
\usepackage[parfill]{parskip}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xr}

\hypersetup{colorlinks=true, linkcolor=black, citecolor=black}
\RequirePackage{lineno}

@@

\def\title{Something occupancy models}
\def\author{Lauren C.\ Ponisio$^{1,2}$, Nicholas Michaud$^1$, Perry de
  Valpine$^1$ (Daniel and Chris?)}

@@

\def\affiliation{
  \begin{enumerate}
  \item Department of Environmental Science, Policy, and Management\\
    University of California, Berkeley\\
    130 Mulford Hall\\
    Berkeley, California, USA\\
    94720\\
  \item Department of Entomology\\
    University of California, Riverside\\
    417 Entomology Bldg.\\
    Riverside, California, USA\\
    92521\\
  \end{enumerate}
}

@@

\begin{abstract}  
  \begin{enumerate}
  \item Something expressing the sentiment: occupancy models are
    everywhere, but model fitting and assessment are extremely
    computationally intensive
  \item Because models are so computationally intensive, users often
    forgo model assessment (determining if a model provides an
    adequate fit to a particular dataset) and model selection
    (choosing the best model out of a set of models) --- methods that
    generally involve simulating from and refitting the model
    iterativly.
  \item Using the open source modeling software NIMBLE, we develop
    combined computational approaches including user-defined and
    automatic blocking of parameters for MCMC, filtering over latent
    states, and customized MCMC samplers for specific parameters to
    improve efficiency. We test these approaches using three
    representative occupancy models of varying levels of complexity
    including a single species model with spatial auto-correlation, a
    single species dynamic (multi-season) model, and a multi-species
    model. We also develop and implement methods for calculating
    calibrated predictive posterior $p$-values to assess model fit and
    cross validation for model selection within NIMBLE.
  \item These computation approaches lead to an improvement in MCMC
    sampling efficiency over, particularly with models including
    random effects. (more results once they are available)
  \item \textit{Implications:} Ours results highlight the need for
    more customizable approaches to MCMC to fit and assess
    hierarchical models in order to ensure occupancy and other
    hierarchical models are accessible to practitioners. By
    implementing MCMC procedures and model assessment and selection
    techniques in open source software, we have made progress toward
    this aim.
  \end{enumerate}
\end{abstract}

@@

Beyond assessing the fit of a model, choosing between models is one of
the most widely used applications of statistics by
practitioners. Though many methods for Bayesian model selection have
been developed \citep{hooten2014guide}, but they, like model
assessment, are computationally intensive. For example,
cross-validation, one of the most fudamental procedures in model
selection, requires iternativly re-fitting the model. A typical need
for model selection arises when a practitioner is choosing whether to
include a specific layer of hierarchy (i.e., random effect). This is
often the case with so called ``multi-species'' occupancy models,
where the occupancy of many species is estimated simultaneously in a
model with a random effect of species \citep[reviewed in,
][]{iknayan2014detecting}. Ecologists are often interested in whether
there is some variability in the response of species to an explanatory
variable such that a random effect of species accounts for that
variability \citep{pacifici2014guidelines}. Currently, the Deviance
Information Criteria (DIC), originally derived to mimic AIC for
Bayesian, non-hierarchical models, is now commonly used by scientists
to evaluate hierarchical models. Though the limitations of DIC for
hierarchical model selection are widely recognized by statisticians
\citep{celeux2006deviance, hooten2014guide}, because it is built into
open-source software such as WinBUGS, it can be used uncritically by
practitioners. Readily available and theoretically sound alternative
methods are thus greatly needed.

@@

We implemented a procedure to calculate calibrated posterior
predictive $p$-values \citep[CPPP, ][]{hjort-etal-06} to assesss the
fit of the model to the data. After the parameters have been fit to
the model, a sample of the posterior is used to simulate data from the
model. A discrepancy measure, which we chose to be the model
likelihood, is then calculated, and the posterior $p$-value is the
number of simulated $p$-values that fall below the observed. To
``calibrate'' the distribution of posterior $p$-values, the MCMC is
rerun on the simulated data to refit the model. If the CPPP $< 0.05$,
the model is rejected as having an adequate fit ot the data
\citep{hjort-etal-06}.

@@

As with the spatail model, we use calibrated posterior predictive
$p$-values \citep{hjort-etal-06} to assesss the fit of the model to
the data.

@@

The last model we analyze is a multi-species, single season occupancy
model examining the effect of wildlife management and habitat
characteristics on bird communities \citep{zipkin2010multi}. The
species-specific coefficients for the effect of basal tree area,
understory foliage and deer management where bound together by a
common distribution with an estimated variance.  For species $i$, we
let $z_{i,j}$ denote its true occupancy state at site $j$. We assumed
that the occupancy of the $i^{\mathrm{th}}$ species at the
$j^{\mathrm{th}}$ site is a Bernoulli random variable $z_{i,j} \sim
\mathrm{Bern}(\psi_{i,j})$.  We then let $x_{i,j,k}$ indicate whether
species $i$ was ($x_{i,j,k}=1$) or was not detected ($x_{i,j,k}=0$) in
the $k^{\mathrm{th}}$ visit to site $j$. We also assumed that
detection was distributed according to be a Bernoulli random variable
such that $x_{i,j,k} \sim \mathrm{Bern}(p_{i}*z_{i,j})$, where $p_{i}$
is the probability that the $i^{\mathrm{th}}$ species was
detected. Both site occupancy and detection were influence by habiat
and survey characteristics \citep{zipkin2010multi}. Specifically,
occurence depended on the study area (CATO, Ind=1, or FCW, Ind=0), the
basal tree area (BA) and the understory foliage cover (UFC). The
species-specific occupancy probabilities are modeled as

@@

We also use calibrated posterior predictive $p$-values
\citep{hjort-etal-06} to assesss the fit of the model to the data.  In
multi-species occupancy models, practitioners are often interested in
determining whether a model including a species random effect for
explanatory variables is a better fit than a model without the random
effect. We implemented a cross-validation procedure for this model
where the detection data for species is left out, the model refitted,
and the fitted model used to predict the occurence of that
species. The predictive error of the model included a random effect of
species is then compared to a model where no species random effects
were included.

@@

\section*{Results}
\label{sec:results}
\subsubsection*{Single species, single season occupancy model with
  spatial auto-correlation}

\subsubsection*{Single species, multi season (dynamic) occupancy
  model}

@@

% \begin{figure}
%   \centering
%   \includegraphics[width=.8\textwidth]{../analysis/changePoint/plotting/networks.pdf}
%   \caption{Assembling hedgerow networks had more changing points
%     (vertical red lines) than non-assembling hedgerows and weedy field
%     margins (a representative sample of non-assembling sites are
%     depicted here). In each network, plants and pollinators are
%     represented by green and yellow circles, respectively, weighted by
%     their degree. Each species has a consistent position in the
%     network across years. In the assembling hedgerows, colored squares
%     in the corner of each network represent the year of
%     assembly. Before hedgerows are installed (when they are still
%     field margins) the year of assembly is zero.}
%   \label{fig:changePoints}
% \end{figure}
% \clearpage

>>>>

\documentclass[12pt]{article} 
\usepackage[pdftex]{graphicx}
\usepackage{natbib} 
\usepackage{color}
\usepackage{amsmath} 
\usepackage{amssymb} 
\usepackage{verbatim}
\usepackage{mathpazo} 
\usepackage{setspace}
\usepackage{multirow}
\usepackage{fullpage}
\usepackage{lscape}
\usepackage{fancyhdr}
\usepackage[normalem]{ulem} 
\usepackage{hyperref}
\usepackage[parfill]{parskip}
\usepackage{graphicx}
\usepackage{rotating}
\usepackage{textcomp}
\hypersetup{colorlinks=true, linkcolor=black, citecolor=black}
\RequirePackage{lineno}

@@

\def\title{Improving the efficiency of occupancy models (something
  more interesting sounding?)}
\def\author{Lauren C.\ Ponisio$^{1,2,3}$, Nicholas Michaud$^1$, Perry de
  Valpine$^1$ (Daniel and Chris?)}

@@

\def\affiliation{
  \begin{enumerate}
  \item Department of Environmental Science, Policy, and Management\\
    University of California, Berkeley\\
    130 Mulford Hall\\
    Berkeley, California, USA\\
    94720\\
  \item Berkeley Institute for Data Science (BIDS)\\
    University of California, Berkeley\\
    190 Doe Library \\
  \item Department of Entomology\\
    University of California, Riverside\\
    417 Entomology Bldg.\\
    Riverside, California, USA\\
    92521\\
  \end{enumerate}
}

@@

\begin{abstract}  
  \begin{enumerate}
  \item Something expressing the sentiment: occupancy models are
    everywhere, but model fitting and assessment are extremely
    computationally intensive
  \item Because models are so computationally intensive, users often
    forgo model assessment (determining if a model provides an
    adequate fit to a particular dataset) and model selection
    (choosing the best model out of a set of models) --- methods that
    generally involve simulating from and refitting the model
    iterativly.
  \item Using the open source modeling software NIMBLE, we develop
    combined computational approaches including user-defined and
    automatic blocking of parameters for MCMC, filtering over latent
    states, and customized MCMC samplers for specific parameters to
    improve efficiency. We test these approaches using three
    representative occupancy models of varying levels of complexity
    including a single species model with spatial auto-correlation, a
    single species dynamic (multi-season) model, and a multi-species
    model. We also develop and implement methods for calculating
    calibrated predictive posterior $p$-values to assess model fit and
    cross validation for model selection within NIMBLE.
  \item These computation approaches lead to an improvement in MCMC
    sampling efficiency over, particularly with models including
    random effects (maybe?). (more results once they are available)
  \item \textit{Implications:} Ours results highlight the need for
    more customizable approaches to MCMC to fit and assess
    hierarchical models in order to ensure occupancy and other
    hierarchical models are accessible to practitioners. By
    implementing MCMC procedures and model assessment and selection
    techniques in open source software, we have made progress toward
    this aim.
  \end{enumerate}
\end{abstract}

@@

Beyond assessing the fit of a model, choosing between models is one of
the most widely used applications of statistics by
practitioners. Though many methods for Bayesian model selection have
been developed \citep{hooten2014guide}, but they, like model
assessment, are computationally intensive. For example,
cross-validation, one of the most fundamental procedures in model
selection, requires iterativly re-fitting the model. A typical need
for model selection arises when a practitioner is choosing whether to
include a specific layer of hierarchy (i.e., random effect). This is
often the case with so called ``multi-species'' occupancy models,
where the occupancy of many species is estimated simultaneously in a
model with a random effect of species \citep[reviewed in,
][]{iknayan2014detecting}. Ecologists are often interested in whether
there is some variability in the response of species to an explanatory
variable such that a random effect of species accounts for that
variability \citep{pacifici2014guidelines}. Currently, the Deviance
Information Criteria (DIC), originally derived to mimic AIC for
Bayesian, non-hierarchical models, is now commonly used by scientists
to evaluate hierarchical models. Though the limitations of DIC for
hierarchical model selection are widely recognized by statisticians
\citep{celeux2006deviance, hooten2014guide}, because it is built into
open-source software such as WinBUGS, it can be used uncritically by
practitioners. Readily available and theoretically sound alternative
methods are thus greatly needed.

@@

We implemented a procedure to calculate calibrated posterior
predictive $p$-values \citep[CPPP, ][]{hjort-etal-06} to assess the
fit of the model to the data. After the parameters have been fit to
the model, a sample of the posterior is used to simulate data from the
model. A discrepancy measure, which we chose to be the model
likelihood, is then calculated, and the posterior $p$-value is the
number of simulated $p$-values that fall below the observed. To
``calibrate'' the distribution of posterior $p$-values, the MCMC is
rerun on the simulated data to refit the model. If the CPPP $< 0.05$,
the model is rejected as having an adequate fit to the data
\citep{hjort-etal-06}.

@@

As with the spatial model, we use calibrated posterior predictive
$p$-values \citep{hjort-etal-06} to assess the fit of the model to
the data.

@@

The last model we analyze is a multi-species, single season occupancy
model examining the effect of wildlife management and habitat
characteristics on bird communities \citep{zipkin2010multi}. The
species-specific coefficients for the effect of basal tree area,
understory foliage and deer management where bound together by a
common distribution with an estimated variance.  For species $i$, we
let $z_{i,j}$ denote its true occupancy state at site $j$. We assumed
that the occupancy of the $i^{\mathrm{th}}$ species at the
$j^{\mathrm{th}}$ site is a Bernoulli random variable $z_{i,j} \sim
\mathrm{Bern}(\psi_{i,j})$.  We then let $x_{i,j,k}$ indicate whether
species $i$ was ($x_{i,j,k}=1$) or was not detected ($x_{i,j,k}=0$) in
the $k^{\mathrm{th}}$ visit to site $j$. We also assumed that
detection was distributed according to be a Bernoulli random variable
such that $x_{i,j,k} \sim \mathrm{Bern}(p_{i}*z_{i,j})$, where $p_{i}$
is the probability that the $i^{\mathrm{th}}$ species was
detected. Both site occupancy and detection were influence by habitat
and survey characteristics \citep{zipkin2010multi}. Specifically,
occurrence depended on the study area (CATO, Ind=1, or FCW, Ind=0), the
basal tree area (BA) and the understory foliage cover (UFC). The
species-specific occupancy probabilities are modeled as

@@

We also use calibrated posterior predictive $p$-values
\citep{hjort-etal-06} to assess the fit of the model to the data.  In
multi-species occupancy models, practitioners are often interested in
determining whether a model including a species random effect for
explanatory variables is a better fit than a model without the random
effect. We implemented a cross-validation procedure for this model
where the detection data for species is left out, the model refitted,
and the fitted model used to predict the occurrence of that
species. The predictive error of the model included a random effect of
species is then compared to a model where no species random effects
were included.

@@

\section*{Results}
\label{sec:results}

\subsubsection*{Multi species, single season occupancy model}
\label{sec:msss-res}

\subsubsection*{Single species, single season occupancy model with
  spatial auto-correlation}
\label{sec:sp-res}

\subsubsection*{Single species, multi season (dynamic) occupancy
  model}
\label{sec:ssms-res}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\textwidth]{../../../saved/SingleSpp-MultiSea/figures/comparisons/SingleSpp-MultiSeaBar.pdf}
  \caption{The mean and minimum efficiency of different MCMC samplers
    for the single species, multi season occupancy model.}
  \label{fig:ss-ms-eff}
\end{figure}
\clearpage

\begin{sidewaysfigure}
  \centering
  \includegraphics[width=.8\textwidth]{../../../saved/SingleSpp-MultiSea/figures/comparisons/SingleSpp-MultiSeaDiffs.pdf}
  \caption{The log ratio of the NIMBLE samplers in comparison to the
    JAGS sampler for the single species, multi season occupancy
    model.}
  \label{fig:ss-ms-eff}
\end{sidewaysfigure}
\clearpage

\begin{figure}
  \centering
  \includegraphics[width=.8\textwidth]{../../../saved/multiSpp-singleSea/figures/comparisons/MultiSpp-SingleSeaBar.pdf}
  \caption{The mean and minimum efficiency of different MCMC samplers
    for the multi species, single season occupancy model.}
  \label{fig:ms-ss-eff}
\end{figure}
\clearpage

\begin{sidewaysfigure}
  \centering
  \includegraphics[width=.8\textwidth]{../../../saved/multiSpp-singleSea/figures/comparisons/MultiSpp-SingleSeaDiffs.pdf}
  \caption{The log ratio of the NIMBLE samplers in comparison to the
    JAGS sampler for the multi species, single season occupancy
    model.}
  \label{fig:ms-ss-eff}
\end{sidewaysfigure}
\clearpage
